---
title: "Getting started with eemr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with eemr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

**eemr** is an R/Rcpp binding for the original *eemCpp* command-line tool. It
implements coherence-based expression module (EEM) search: given an
expression matrix and a collection of candidate gene sets (e.g. from MSigDB),
it finds, for each gene set, the largest subset of genes whose expression is
coherent across samples (an "expression module"), and assigns that module a
significance p-value based on a randomization test. It's conceptually similar
to gene set enrichment analysis, but the "signal" being tested is coherence
of expression rather than differential expression, and only the coherent
subset of each gene set is retained.

The whole pipeline runs in memory from R -- no intermediate files, no
external CLI binary.

```{r setup}
library(eemr)
```

## Input data

`eem_search()` takes two inputs, already loaded into R: `expr`, a numeric
matrix, and `geneSets`, a named list of character vectors. Both are commonly
distributed as plain-text files, so this package bundles a small real example
of each and provides `read_gmt()` to read the gene set file.

### Expression matrix (`.tab`)

A tab-delimited text file: the first line is a header giving the sample
names (with an empty first field), and each following line is one gene --
its name, then one tab-separated numeric value per sample:

```
	sample1	sample2	sample3
GENE1	4.68	5.37	13.16
GENE2	4.72	10.36	12.11
```

This is exactly the format read by `read.table(file, header = TRUE, row.names = 1, sep = "\t", check.names = FALSE)`, which is how the bundled `test.tab`
is loaded below. `check.names = FALSE` keeps sample names that aren't valid R
identifiers (e.g. starting with a digit) unchanged. The result is coerced to
a plain numeric `matrix` with `as.matrix()` because `eem_search()` requires a
matrix, not a data.frame.

### Gene sets (`.gmt`)

A tab-delimited GMT file, the format used by MSigDB/GSEA: one gene set per
line, with the gene set id, a description (unused here, but required by the
format -- `"na"` if you don't have one), and then one gene id per remaining
field:

```
GENE_SET_1	na	GENE1	GENE2	GENE3
GENE_SET_2	na	GENE4	GENE5	GENE6	GENE7
```

`read_gmt(file)` reads this into a named list, e.g.
`list(GENE_SET_1 = c("GENE1", "GENE2", "GENE3"), ...)`, suitable for
`eem_search()`'s `geneSets` argument.

This package bundles a small real dataset (1000 genes x 118 breast cancer
samples, and 100 candidate GMT gene sets) for examples and testing:

```{r load-data}
expr_file <- system.file("extdata", "test.tab", package = "eemr")
gmt_file  <- system.file("extdata", "test.gmt", package = "eemr")

expr <- as.matrix(read.table(expr_file, header = TRUE, row.names = 1,
                              sep = "\t", check.names = FALSE))
geneSets <- read_gmt(gmt_file)

dim(expr)
length(geneSets)
```

## Running the search

```{r eem-search, results = "hide"}
res <- eem_search(expr, geneSets)
```

The result is a data.frame with one row per gene set that passed the search,
sorted by decreasing `pvalue` (more significant first):

```{r show-results}
head(res[, c("id", "pvalue", "pvalue1", "pvalue2",
             "nSeedGenes", "nModuleGenes")], 10)
```

* `pvalue1` is an approximate, hypergeometric-based p-value used to screen
  candidate gene sets cheaply.
* `pvalue2` is the accurate, permutation-based p-value, calculated only for
  gene sets that pass the `pvalue1Cutoff` screen (`-1` means it wasn't
  calculated).
* `pvalue` is the one actually reported (`pvalue2` when available, otherwise
  `pvalue1`).
* All three are on the -log10 scale, so larger values are more significant.
* `seedGenes`/`moduleGenes` list the gene set's original genes and the
  coherent subset found by the search (`;`-separated).

Key parameters (see `?eem_search`):

* `relativeRadius`: target fraction of all genes that fall within the
  module search radius (default `0.05`).
* `pvalue1Cutoff`: -log10 cutoff used to screen candidates before the
  accurate p-value is calculated (default `-log10(0.05)`); use `-1` to
  disable screening.
* `itr`: number of permutations for the accurate p-value (default `300`).

## Module activity matrix

Once you have modules, `module_activity()` turns them into a module x sample
matrix, useful for downstream analyses (clustering samples, correlating with
clinical variables, etc.). For each module it averages the (row-standardized,
by default) expression of its `moduleGenes` across samples:

```{r module-activity}
activity <- module_activity(expr, res)
dim(activity)
activity[1:5, 1:5]
```

By default only modules with `pvalue >= 6` (-log10 scale, i.e. p < 1e-6) are
included; pass `pvalueCutoff = NULL` to include every module in `res`, or any
other -log10 threshold:

```{r module-activity-cutoff}
nrow(module_activity(expr, res, pvalueCutoff = NULL))
nrow(module_activity(expr, res, pvalueCutoff = 10))
```

A quick look at how the top modules' activity varies across samples:

```{r heatmap, fig.width = 7, fig.height = 5}
heatmap(activity, scale = "none", margins = c(2, 12),
        labCol = FALSE, main = "Module activity")
```
